Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-01-31 parallel-tractability-J-v9/parallelTractabilityDatalog.tex
Section 4 and small adjustments done

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
\end{enumerate}
For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are two nodes,
we say that the node $v_1$ is the parent (node) of $v_2$ and the node $v_2$ is the child (node)
of $v_1$. Further, $\forall H,B_1,...,B_n\in V$, $\mathcal{G}$ satisfies the following conditions:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $H$ has a parent node or a child node;
  \item if $H$ has the in-degree of $0$, then $H$ is an original fact of $P$;
  \item $B_1,...,B_n\rightarrow H\in P^*$ is satisfied if $e(B_1, H),...,e(B_n, H)\in E$ and
  $B_1,...,B_n$ are \d{all} the parents of $H$.
\end{enumerate}
\end{definition}

@@

\begin{example}\label{exp:mg}
Consider a DHL($\circ$) ontology $\mathcal{O}_{ex_1}$ where the TBox is
$\{\exists R.A\sqsubseteq A\}$, the RBox is $\{S\circ R\sqsubseteq R\}$ and
the ABox is $\{A(b),R(a_1,b),S(a_i,a_{i-1})\}$ for $2\leq i\leq k$ and
$k$ is an integer greater than $2$.
The corresponding datalog program of this ontology is $P_{ex_1}=\langle R, \textbf{I}\rangle$
where $\textbf{I}$ contains all the assertions in the ABox
and $R$ contains the two rules `$R(x,y),A(y)\rightarrow A(x)$'
and `$S(x,y),R(y,z)\rightarrow R(x,z)$'.
The graph in Figure~\ref{fig:mg} is a materialization graph with respect to $P_{ex_1}$,
denoted by $\mathcal{G}_{ex_1}$.
The nodes with in-degree 0 are the original facts in $\textbf{I}$;
each of the other nodes corresponds to a ground instantiation of some rule.
For example, the node $A(a_k)$ corresponds to the ground rule instantiation
`$R(a_k,b),A(b)\rightarrow A(a_k)$'.
The size of this materialization graph is the number of nodes, that is $3k$.
The depth of $\mathcal{G}_{ex_1}$ is $k$.
\end{example}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig-mg.eps}
\caption{An example of a materialization graph.}
\label{fig:mg}
\end{center}
\end{figure}

@@

In this part, we propose a parallel algorithm
(denoted by Algorithm~$\mathsf{A}_{bsc}$) that constructs a materialization graph for a given datalog program.
Our computational model is a PRAM (parallel, random-access machine) that is mostly used to analyze parallel
complexity. This model allows us to introduce \emph{the parallel assumption}:
for any datalog program $P=\langle R, \textbf{I}\rangle$,
any substitution of some atom and any rule instance in $P^*$ can be mapped to a unique memory
location; further, a one-to-one relation can be established between processors and rule instances.
Under this assumption, a processor can check the applicability of its corresponding
rule instance and access the state of an atom occurring in this rule instance in constant time.
Algorithm~$\mathsf{A}_{bsc}$ is then given as follows.\\

\noindent\texttt{Algorithm~$\mathsf{A}_{bsc}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$,
the algorithm returns a materialization graph $\mathcal{G}$ of $P$.
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.
Initially $\mathcal{G}$ is empty. The following three steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg1:addFacts}
\item For each rule instantiation $B_1,...,B_n\rightarrow H$, if the body atoms are all
    in $\mathcal{G}$ while $H$ is not in $\mathcal{G}$,
    the corresponding processor adds $H$ to $\mathcal{G}$ and creates edges pointing
    from $B_1,...,B_n$ to $H$.\label{alg1:updateG}
\item If no processor can add more nodes and edges to $\mathcal{G}$,
  terminate, otherwise continue with \ref{alg1:updateG}.\label{alg1:halt}\hfill$\Box$
\end{enumerate}

\begin{example}
We consider the datalog program $P_{ex_1}$ in Example~\ref{exp:mg} again,
and perform Algorithm~$\mathsf{A}_{bsc}$ on it.
Initially, all the facts ($A(b),R(a_1,b),S(a_2,a_1),...,S(a_{k},a_{k-1})$) are added to the
result $\mathcal{G}_{ex_1}$ (\ref{alg1:addFacts}).
Then in different iterations of \ref{alg1:updateG}, the remaining nodes are added to
$\mathcal{G}_{ex_1}$ by different processors.
For example a processor $p$ is allocated a rule instantiation `$R(a_2,b),A(b)\rightarrow A(a_2)$'.
Then, processor $p$ adds $A(a_2)$ to $\mathcal{G}_{ex_1}$ after it checks that
$A(b)$ and $R(a_2,b)$ are in $\mathcal{G}_{ex_1}$.
Algorithm~$\mathsf{A}_{bsc}$ halts when $A(a_k)$ has been added to $\mathcal{G}_{ex_1}$ (\ref{alg1:halt}).
\end{example}

Lemma~\ref{lemma:a1} shows the correctness of Algorithm~$\mathsf{A}_{bsc}$ and that, for any datalog program $P$,
Algorithm~$\mathsf{A}_{bsc}$ always constructs a materialization graph with the minimum depth among all the
materialization graphs of $P$. The detailed proofs of Lemma~\ref{lemma:a1} and other lemmas and theorems can be found
in the appendix.

\begin{lemma}
\label{lemma:a1}
Given a datalog program $P=\langle R, \textbf{I}\rangle$, we have
\begin{enumerate}[leftmargin=4ex]
\item Algorithm~$\mathsf{A}_{bsc}$ halts and returns a materialization graph $\mathcal{G}$ of $P$;
\item $\mathcal{G}$ has the the minimum depth among all the materialization graphs of $P$.
\end{enumerate}
\end{lemma}

\begin{proof}[Proof sketch] This lemma can be proved by performing
an induction on $T_R^{\omega}(\textbf{I})$.
The stage (see the related contents in Section~\ref{sec:background}) of $P$
is the lower-bound of the depth of the materialization graphs. Based on the previous induction,
one can further check that, for the materialization graph $\mathcal{G}$ constructed by Algorithm~$\mathsf{A}_{bsc}$,
its depth equals the depth of the stage.
\end{proof}

We now discuss the parallel complexity of Algorithm~$\mathsf{A}_{bsc}$.
Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$.
Let $e$, $v$ and $r$ represent the maximum arity of any EDB predicate, the maximum number of variables
in any datalog rule, and the number of datalog rules, respectively.
We then have that the number of
constants is at most $|\textbf{I}|e$, and the number of all possible rule instances in $P^*$
is at most $r(|\textbf{I}|e)^v$.
Note that $e, v$ and $r$ depend only on the rule set $R$ and not on the fact set $\textbf{I}$.
Thus, the memory space for storing the atoms and the rule instances is polynomial in the size of $\textbf{I}$.
This also means that the number of processors is polynomially bounded.
The computing time of \ref{alg1:addFacts} and \ref{alg1:halt} occupy constant time (denoted by $c_1$) because of parallelism.
Since Algorithm~$\mathsf{A}_{bsc}$ works under the parallel assumption, one iteration of
\ref{alg1:updateG} also costs constant time (denoted by $c_2$). Thus,
the whole computing time of Algorithm~$\mathsf{A}_{bsc}$ turns out to be $c_1+l\cdot c_2$
where $l$ denotes the number of iterations of \ref{alg1:updateG}.

An \texttt{NC} algorithm should meet two requirements: first,
it works on a polynomial number of processors; second, it halts in poly-logarithmic time.
As discussed above, Algorithm~$\mathsf{A}_{bsc}$ meets the first requirement.
If we want to restrict Algorithm~$\mathsf{A}_{bsc}$ to be an \texttt{NC} algorithm,
we can make the number of iterations of \ref{alg1:updateG} to be poly-logarithmically bounded.
We use the symbol $\psi$ to denote a poly-logarithmically bounded function.
For any datalog program, if we restrict the number of iterations of
\ref{alg1:updateG} to be bounded by $\psi(|\textbf{I}|)$,
the computing time of Algorithm~$\mathsf{A}_{bsc}$
is $c_1+\psi(|\textbf{I}|)\cdot c_2$. With this restriction, Algorithm~$\mathsf{A}_{bsc}$ is an \texttt{NC} algorithm denoted by
$\mathsf{A}_{bsc}^{\psi}$.

Based on $\mathsf{A}_{bsc}^{\psi}$, we can identify a class of datalog programs
$\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$ such that all the datalog programs in it can be handled
by $\mathsf{A}_{bsc}^{\psi}$.
It is obvious that $\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$ is a \texttt{PTD} class.
We use the following theorem to further show that this class can be captured based on
the materialization graphs of the datalog programs in $\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$.

\begin{theorem}\label{theorem:a1}
For any datalog program $P$, $P\in\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$ iff $P$ has a
materialization graph whose depth is upper-bounded by $\psi(|P|)$.
\end{theorem}

@@

Consider Example~\ref{exp:mg} again. Let the integer $k$ be a variable. We can
get a class of datalog programs, denoted by $\mathbb{P}_{ex_1}$, where the rule set
is $\{R(x,y),A(y)\rightarrow A(x), S(x,y),R(y,z)\rightarrow R(x,z)\}$ and the fact
set varies according to $k$. The algorithm $\mathsf{A}_{bsc}^{\psi}$ is restricted in the sense that it cannot even work on the rather simple datalog program class $\mathbb{P}_{ex_1}$.
Let $\mathcal{G}_{ex_1}$ be some materialization graph corresponding
to some datalog program in $\mathbb{P}_{ex_1}$. It can be checked that
\texttt{depth}($\mathcal{G}_{ex_1}$)$=k$ for some $k$.
This means that the depths of the materialization graphs are linearly bounded by $k$.
On the other hand, the sizes of the datalog programs in $\mathbb{P}_{ex_1}$ are polynomial in $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{bsc}^{\psi}$ terminates without constructing a materialization
graph for each datalog program in $\mathbb{P}_{ex_1}$.
However, there indeed exists an \texttt{NC} algorithm that can handle $\mathbb{P}_{ex_1}$.
We discuss this in the next part.

\subsection{Optimizing Algorithm~$\mathsf{A}_{bsc}$ via Single-Way Derivability}

In this part, we optimize Algorithm~$\mathsf{A}_{bsc}$ such that $P_{ex_1}$
can be handled. Based on the optimized variant of Algorithm~$\mathsf{A}_{bsc}$,
we can identify another \texttt{PTD} class.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig-general.eps}
\caption{A partial materialization graph}
\label{fig:general}
\end{center}
\end{figure}

\begin{example}\label{exp:general}
Consider the  snapshot of Algorithm~$\mathsf{A}_{bsc}$ in
Figure~\ref{fig:general}. A materialization graph $\mathcal{G}$
is being constructed for some datalog program $\langle R, \textbf{I}\rangle$.
The nodes in the dashed box denote the ones that have been added to $\mathcal{G}$.
In this snapshot, $v_0$ has been newly added
to $\mathcal{G}$ in the $\ell^{th}$ ($\ell\geq 1$) iteration.
Each of the nodes $v_i$ ($1\leq i\leq k$) has at most one parent node not in $\mathcal{G}$,
while $v'$ has two parent nodes not in $\mathcal{G}$.
All of the nodes $v_i$ ($1\leq i\leq k$) and $v'$ would be added to $\mathcal{G}$
afterwards.
\end{example}

In Example~\ref{exp:general}, $v_k$ would be added to $\mathcal{G}$ after \emph{at least} $k$
iterations by performing Algorithm~$\mathsf{A}_{bsc}$.
We can check that each edge $(v_{i-1},v_i)$ ($1\leq i\leq k$) in this
stage adheres to the condition:
\emph{if the parent node $v_{i-1}$ is derivable, the child node $v_i$ is derivable} (this
is because for a node $v_i$, the node $v_{i-1}$ is the only parent node that has not been added to $\mathcal{G}$).
We call this condition the \emph{single-way derivability condition}, which intuitively says that the derivability of a
node only depends on one of its parent nodes.
Observe that node $v_k$ is reachable from node $v_0$ through the path $\tau=(v_0,v_1,\ldots,v_k)$ where
each edge $(v_{i-1},v_i)$ ($1\leq i\leq k$) satisfies the single-way derivability condition.
We call such a path a \emph{single-way derivable} (SWD) path.
Further, the starting node $v_0$ in $\tau$ has been added to $\mathcal{G}$; in other words,
$v_0$ is derivable. Thus, all of the nodes $v_i$ ($1\leq i\leq k$) can be added to
$\mathcal{G}$ immediately.
Based on the above idea, we optimize Algorithm~$\mathsf{A}_{bsc}$ by checking whether such an SWD path
exits for some node $v$. If there is an SWD path for node $v$, this node can be added to $\mathcal{G}$
due to the single-way derivability condition.

@@

We next discuss how to determine the existence of SWD paths. Note that, SWD paths require us to
describe the reachability between two nodes. To this extent, we use a
binary transitive relation $\texttt{rch} \subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$,
e.g., \texttt{rch}$(v_1,v_2)$ means that $v_2$ is reachable from $v_1$.
In each iteration of \ref{alg1:updateG} in Algorithm~$\mathsf{A}_{bsc}$, we further compute a \texttt{rch}
relation (denoted by $S_{\textit{\tiny rch}}$) by performing the following process:

\begin{description}[leftmargin=4ex]
\item[(\textbf{\dag})] \emph{For each rule instantiation of the form $B_1,..,B_i,..,B_n\rightarrow H$
such that $H$ has not been added to $\mathcal{G}$}:
\begin{enumerate}[leftmargin=0ex]
\item \emph{if the all body atoms $B_1,...,B_n$ have been added to $\mathcal{G}$, put \texttt{rch}$(B_1,H),...,$ \texttt{rch}$(B_n,H)$ in $S_{\textit{\tiny rch}}$};
\item \emph{if $B_i$ is the only node in the body that has not been added to $\mathcal{G}$,
    put \texttt{rch}$(B_i,H)$ in $S_{\textit{\tiny rch}}$}.\hfill$\Box$
\end{enumerate}
\end{description}

@@

\noindent\texttt{Algorithm~$\mathsf{OPT}$}. The algorithm requires two inputs:
a datalog program $P=\langle R, \textbf{I}\rangle$ and a (partial) materialization graph $\mathcal{G}$ that is
being constructed from $P$. The following steps are performed:
\begin{enumerate}[leftmargin=6ex,label=(\textbf{\roman*})]
\item Compute a \texttt{rch} relation $S_{\!\textit{\tiny rch}}$ by following the above process (see $(\textbf{\dag})$).\label{rch}
\item Compute the transitive closure $S^*_{\!\textit{\tiny rch}}$ of $S_{\!\textit{\tiny rch}}$.\label{transClos}
\item Update $\mathcal{G}$ as follows: for any \texttt{rch}$(B_i,H)\in S_{\!\textit{\tiny rch}}$
that corresponds to `$B_1,..,B_i,..$ $,B_n\rightarrow H$'
    and there exists a node $B'$ such that \texttt{rch}$(B',B_i)\in S^*_{\!\textit{\tiny rch}}$ and $B'$ is
    in $\mathcal{G}$; if $H$ is not in $\mathcal{G}$ or $H$ is in $\mathcal{G}$ but has no parent pointing
    to it, add $H$ and $B_i$ (if $B_i$ is not in $\mathcal{G}$) to $\mathcal{G}$, and create the edges
    $e(B_1, H),...,e(B_n, H)$ in $\mathcal{G}$. Do nothing for other statements
    \texttt{rch}$(B_j,H)\in S_{\!\textit{\tiny rch}}$.\label{updateG}\hfill$\Box$
\end{enumerate}

It is well known that there is an \texttt{NC} algorithm for computing the
transitive closure \cite{Allender07}.
Based on this result and Algorithm~$\mathsf{OPT}$, we propose an optimized variant of Algorithm~$\mathsf{A}_{bsc}$:\\

>>>>

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
\end{enumerate}
For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are two nodes,
we say that the node $v_1$ is the parent (node) of $v_2$ and the node $v_2$ is the child (node)
of $v_1$. Further, $\forall H,B_1,\ldots,B_n\in V$, $\mathcal{G}$ satisfies the following conditions:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $H$ has a parent node or a child node;
  \item if $H$ has the in-degree of $0$, then $H$ is an original fact of $P$;
  \item $B_1,\ldots,B_n\rightarrow H\in P^*$ is satisfied if $e(B_1, H),\ldots,e(B_n, H)\in E$ and
  $B_1,\ldots,B_n$ are \d{all} the parents of $H$.
\end{enumerate}
\end{definition}

@@

\begin{example}\label{exp:mg}
Consider a DHL($\circ$) ontology $\mathcal{O}_{\text{ex}_1}$ where the TBox is
$\{\exists R.A\sqsubseteq A\}$, the RBox is $\{S\circ R\sqsubseteq R\}$ and
the ABox is $\{A(b),R(a_1,b),S(a_i,a_{i-1})\}$ for $2\leq i\leq k$ and
$k$ an integer greater than $2$.
The corresponding datalog program of this ontology is $P_{\text{ex}_1}=\langle R, \textbf{I}\rangle$
where $\textbf{I}$ contains all the assertions in the ABox
and $R$ contains the two rules `$R(x,y),A(y)\rightarrow A(x)$'
and `$S(x,y),R(y,z)\rightarrow R(x,z)$'.
The graph in Figure~\ref{fig:mg} is a materialization graph with respect to $P_{\text{ex}_1}$,
denoted by $\mathcal{G}_{\text{ex}_1}$.
The nodes with in-degree 0 are the original facts in $\textbf{I}$;
each of the other nodes corresponds to a ground instantiation of some rule.
For example, the node $A(a_k)$ corresponds to the ground rule instantiation
`$R(a_k,b),A(b)\rightarrow A(a_k)$'.
The size of this materialization graph is the number of nodes, that is $3k$.
The depth of $\mathcal{G}_{\text{ex}_1}$ is $k$.
\end{example}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig-mg.pdf}
\caption{An example materialization graph}
\label{fig:mg}
\end{center}
\end{figure}

@@

In this part, we propose a parallel algorithm
(denoted by Algorithm~$\mathsf{A}_{\text{bsc}}$) that constructs a materialization graph for a given datalog program.
Our computational model is a PRAM (parallel, random-access machine) that is mostly used to analyze parallel
complexity. This model allows us to introduce \emph{the parallel assumption}:
for any datalog program $P=\langle R, \textbf{I}\rangle$,
any substitution of some atom and any rule instance in $P^*$ can be mapped to a unique memory
location; further, a one-to-one relation can be established between processors and rule instances.
Under this assumption, a processor can check the applicability of its corresponding
rule instance and access the state of an atom occurring in this rule instance in constant time.
Algorithm~$\mathsf{A}_{\text{bsc}}$ is then given as follows.\\

\noindent\texttt{Algorithm~$\mathsf{A}_{\text{bsc}}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$,
the algorithm returns a materialization graph $\mathcal{G}$ of $P$.
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.
Initially $\mathcal{G}$ is empty. The following three steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg1:addFacts}
\item For each rule instantiation $B_1,\ldots,B_n\rightarrow H$, if the body atoms are all
    in $\mathcal{G}$ while $H$ is not in $\mathcal{G}$,
    the corresponding processor adds $H$ to $\mathcal{G}$ and creates edges pointing
    from $B_1,\ldots,B_n$ to $H$.\label{alg1:updateG}
\item If no processor can add more nodes and edges to $\mathcal{G}$,
  terminate, otherwise continue with \ref{alg1:updateG}.\label{alg1:halt}\hfill$\Box$
\end{enumerate}

\begin{example}
We consider the datalog program $P_{\text{ex}_1}$ in Example~\ref{exp:mg} again,
and perform Algorithm~$\mathsf{A}_{\text{bsc}}$ on it.
Initially, all the facts ($A(b),R(a_1,b),S(a_2,a_1),\ldots,S(a_{k},a_{k-1})$) are added to the
result $\mathcal{G}_{\text{ex}_1}$ (\ref{alg1:addFacts}).
Then in different iterations of \ref{alg1:updateG}, the remaining nodes are added to
$\mathcal{G}_{\text{ex}_1}$ by different processors.
For example a processor $p$ is allocated a rule instantiation `$R(a_2,b),A(b)\rightarrow A(a_2)$'.
Then, processor $p$ adds $A(a_2)$ to $\mathcal{G}_{\text{ex}_1}$ after it checks that
$A(b)$ and $R(a_2,b)$ are in $\mathcal{G}_{\text{ex}_1}$.
Algorithm~$\mathsf{A}_{\text{bsc}}$ halts when $A(a_k)$ has been added to $\mathcal{G}_{\text{ex}_1}$ (\ref{alg1:halt}).
\end{example}

Lemma~\ref{lemma:a1} shows the correctness of Algorithm~$\mathsf{A}_{\text{bsc}}$ and that, for any datalog program $P$,
Algorithm~$\mathsf{A}_{\text{bsc}}$ always constructs a materialization graph with the minimum depth among all the
materialization graphs of $P$. The detailed proofs of Lemma~\ref{lemma:a1} and other lemmas and theorems can be found
in the appendix.

\begin{lemma}
\label{lemma:a1}
Given a datalog program $P=\langle R, \textbf{I}\rangle$, we have
\begin{enumerate}[leftmargin=4ex]
\item Algorithm~$\mathsf{A}_{\text{bsc}}$ halts and returns a materialization graph $\mathcal{G}$ of $P$;
\item $\mathcal{G}$ has the the minimum depth among all the materialization graphs of $P$.
\end{enumerate}
\end{lemma}

\begin{proof}[Proof sketch] This lemma can be proved by performing
an induction on $T_R^{\omega}(\textbf{I})$.
The stage (see the related contents in Section~\ref{sec:background}) of $P$
is the lower-bound of the depth of the materialization graphs. Based on the previous induction,
one can further check that, for the materialization graph $\mathcal{G}$ constructed by Algorithm~$\mathsf{A}_{\text{bsc}}$,
its depth equals the depth of the stage.
\end{proof}

We now discuss the parallel complexity of Algorithm~$\mathsf{A}_{\text{bsc}}$.
Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$.
Let $e$, $v$ and $r$ represent the maximum arity of any predicate in $\textbf{I}$, the maximum number of variables
in any datalog rule, and the number of datalog rules respectively.
We then have that the number of
constants is at most $|\textbf{I}|e$, and the number of all possible rule instances in $P^*$
is at most $r(|\textbf{I}|e)^v$.
Note that $e, v$ and $r$ depend only on the rule set $R$ and not on the fact set $\textbf{I}$.
Thus, the memory space for storing the atoms and the rule instances is polynomial in the size of $\textbf{I}$.
This also means that the number of processors is polynomially bounded.
The computing time of \ref{alg1:addFacts} and \ref{alg1:halt} occupy constant time (denoted by $c_1$) because of parallelism.
Since Algorithm~$\mathsf{A}_{\text{bsc}}$ works under the parallel assumption, one iteration of
\ref{alg1:updateG} also costs constant time (denoted by $c_2$). Thus,
the whole computing time of Algorithm~$\mathsf{A}_{\text{bsc}}$ turns out to be $c_1+\ell\cdot c_2$
where $\ell$ denotes the number of iterations of \ref{alg1:updateG}.

An \texttt{NC} algorithm should meet two requirements: first,
it works on a polynomial number of processors; second, it halts in poly-logarithmic time.
As discussed above, Algorithm~$\mathsf{A}_{\text{bsc}}$ meets the first requirement.
If we want to restrict Algorithm~$\mathsf{A}_{\text{bsc}}$ to be an \texttt{NC} algorithm,
we can make the number of iterations of \ref{alg1:updateG} to be poly-logarithmically bounded.
We use the symbol $\psi$ to denote a poly-logarithmically bounded function.
For any datalog program, if we restrict the number of iterations of
\ref{alg1:updateG} to be bounded by $\psi(|\textbf{I}|)$,
the computing time of Algorithm~$\mathsf{A}_{\text{bsc}}$
is $c_1+\psi(|\textbf{I}|)\cdot c_2$. With this restriction, Algorithm~$\mathsf{A}_{\text{bsc}}$ is an \texttt{NC} algorithm denoted by
$\mathsf{A}_{\text{bsc}}^{\psi}$.

Based on $\mathsf{A}_{\text{bsc}}^{\psi}$, we can identify a class of datalog programs
$\mathcal{D}_{\mathsf{A}_{\text{bsc}}^{\psi}}$ such that all the datalog programs in it can be handled
by $\mathsf{A}_{\text{bsc}}^{\psi}$.
It is obvious that $\mathcal{D}_{\mathsf{A}_{\text{bsc}}^{\psi}}$ is a \texttt{PTD} class.
We use the following theorem to further show that this class can be captured based on
the materialization graphs of the datalog programs in $\mathcal{D}_{\mathsf{A}_{\text{bsc}}^{\psi}}$.

\begin{theorem}\label{theorem:a1}
For any datalog program $P$, $P\in\mathcal{D}_{\mathsf{A}_{\text{bsc}}^{\psi}}$ iff $P$ has a
materialization graph whose depth is upper-bounded by $\psi(|P|)$.
\end{theorem}

@@

Consider Example~\ref{exp:mg} again. Let the integer $k$ be a variable. We can
get a class of datalog programs, denoted by $\mathbb{P}_{\text{ex}_1}$, where the rule set
is $\{R(x,y),A(y)\rightarrow A(x), S(x,y),R(y,z)\rightarrow R(x,z)\}$ and the fact
set varies according to $k$. The algorithm $\mathsf{A}_{\text{bsc}}^{\psi}$ is restricted in the sense that it cannot even work on the rather simple datalog program class $\mathbb{P}_{\text{ex}_1}$.
Let $\mathcal{G}_{\text{ex}_1}$ be some materialization graph corresponding
to some datalog program in $\mathbb{P}_{\text{ex}_1}$. It can be checked that
\texttt{depth}($\mathcal{G}_{\text{ex}_1}$)$=k$ for some $k$.
This means that the depths of the materialization graphs are linearly bounded by $k$.
On the other hand, the sizes of the datalog programs in $\mathbb{P}_{\text{ex}_1}$ are polynomial in $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{\text{bsc}}^{\psi}$ terminates without constructing a materialization
graph for each datalog program in $\mathbb{P}_{\text{ex}_1}$.
However, there indeed exists an \texttt{NC} algorithm that can handle $\mathbb{P}_{\text{ex}_1}$.
We discuss this in the next part.

\subsection{Optimizing Algorithm~$\mathsf{A}_{\text{bsc}}$ via Single-Way Derivability}

In this part, we optimize Algorithm~$\mathsf{A}_{\text{bsc}}$ such that $P_{\text{ex}_1}$
can be handled. Based on the optimized variant of Algorithm~$\mathsf{A}_{\text{bsc}}$,
we can identify another \texttt{PTD} class.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig-general.pdf}
\caption{A partial materialization graph}
\label{fig:general}
\end{center}
\end{figure}

\begin{example}\label{exp:general}
Consider the  snapshot of Algorithm~$\mathsf{A}_{\text{bsc}}$ in
Figure~\ref{fig:general}. A materialization graph $\mathcal{G}$
is being constructed for some datalog program $\langle R, \textbf{I}\rangle$.
The nodes in the dashed box denote the ones that have been added to $\mathcal{G}$.
In this snapshot, $v_0$ has been newly added
to $\mathcal{G}$ in the $\ell^{th}$ ($\ell\geq 1$) iteration.
Each of the nodes $v_i$ ($1\leq i\leq k$) has at most one parent node not in $\mathcal{G}$,
while $v'$ has two parent nodes not in $\mathcal{G}$.
All of the nodes $v_i$ ($1\leq i\leq k$) and $v'$ would be added to $\mathcal{G}$
afterwards.
\end{example}

In Example~\ref{exp:general}, $v_k$ would be added to $\mathcal{G}$ after \emph{at least} $k$
iterations by performing Algorithm~$\mathsf{A}_{\text{bsc}}$.
We can check that each edge $(v_{i-1},v_i)$ ($1\leq i\leq k$) in this
stage adheres to the condition:
\emph{if the parent node $v_{i-1}$ is derivable, the child node $v_i$ is derivable} (this
is because for a node $v_i$, the node $v_{i-1}$ is the only parent node that has not been added to $\mathcal{G}$).
We call this condition the \emph{single-way derivability condition}, which intuitively says that the derivability of a
node only depends on one of its parent nodes.
Observe that node $v_k$ is reachable from node $v_0$ through the path $\tau=(v_0,v_1,\ldots,v_k)$ where
each edge $(v_{i-1},v_i)$ ($1\leq i\leq k$) satisfies the single-way derivability condition.
We call such a path a \emph{single-way derivable} (SWD) path.
Further, the starting node $v_0$ in $\tau$ has been added to $\mathcal{G}$; in other words,
$v_0$ is derivable. Thus, all of the nodes $v_i$ ($1\leq i\leq k$) can be added to
$\mathcal{G}$ immediately.
Based on the above idea, we optimize Algorithm~$\mathsf{A}_{\text{bsc}}$ by checking whether such an SWD path
exits for some node $v$. If there is an SWD path for node $v$, this node can be added to $\mathcal{G}$
due to the single-way derivability condition.

@@

We next discuss how to determine the existence of SWD paths. Note that, SWD paths require us to
describe the reachability between two nodes. To this extent, we use a
binary transitive relation $\texttt{rch} \subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$,
e.g., \texttt{rch}$(v_1,v_2)$ means that $v_2$ is reachable from $v_1$.
In each iteration of \ref{alg1:updateG} in Algorithm~$\mathsf{A}_{\text{bsc}}$, we further compute a \texttt{rch}
relation (denoted by $S_{\textit{\tiny rch}}$) by performing the following process:

\begin{description}[leftmargin=4ex]
\item[(\textbf{\dag})] \emph{For each rule instantiation of the form $B_1,..,B_i,..,B_n\rightarrow H$
such that $H$ has not been added to $\mathcal{G}$}:
\begin{enumerate}[leftmargin=0ex]
\item \emph{if the all body atoms $B_1,\ldots,B_n$ have been added to $\mathcal{G}$, put \texttt{rch}$(B_1,H),\ldots,$ \texttt{rch}$(B_n,H)$ in $S_{\textit{\tiny rch}}$};
\item \emph{if $B_i$ is the only node in the body that has not been added to $\mathcal{G}$,
    put \texttt{rch}$(B_i,H)$ in $S_{\textit{\tiny rch}}$}.\hfill$\Box$
\end{enumerate}
\end{description}

@@

\noindent\texttt{Algorithm~$\mathsf{OPT}$}. The algorithm requires two inputs:
a datalog program $P=\langle R, \textbf{I}\rangle$ and a (partial) materialization graph $\mathcal{G}$ that is
being constructed from $P$. The following steps are performed:
\begin{enumerate}[leftmargin=6ex,label=(\textbf{\roman*})]
\item Compute the \texttt{rch} relation $S_{\!\textit{\tiny rch}}$ by following the above process (see $(\textbf{\dag})$).\label{rch}
\item Compute the transitive closure $S^*_{\!\textit{\tiny rch}}$ of $S_{\!\textit{\tiny rch}}$.\label{transClos}
\item Update $\mathcal{G}$ as follows: for any \texttt{rch}$(B_i,H)\in S_{\!\textit{\tiny rch}}$
that corresponds to `$B_1,..,B_i,..$ $,B_n\rightarrow H$'
    and there exists a node $B'$ such that \texttt{rch}$(B',B_i)\in S^*_{\!\textit{\tiny rch}}$ and $B'$ is
    in $\mathcal{G}$; if $H$ is not in $\mathcal{G}$ or $H$ is in $\mathcal{G}$ but has no parent pointing
    to it, add $H$ and $B_i$ (if $B_i$ is not in $\mathcal{G}$) to $\mathcal{G}$, and create the edges
    $e(B_1, H), \ldots, e(B_n, H)$ in $\mathcal{G}$. Do nothing for other statements
    \texttt{rch}$(B_j,H)\in S_{\!\textit{\tiny rch}}$.\label{updateG}\hfill$\Box$
\end{enumerate}

It is well known that there is an \texttt{NC} algorithm for computing the
transitive closure \cite{Allender07}.
Based on this result and Algorithm~$\mathsf{OPT}$, we propose an optimized variant of Algorithm~$\mathsf{A}_{\text{bsc}}$:\\
